### 2014-11-08 coordination.tex
More updates.

In order to allow different scheduling strategies, we introduce the concept of
\emph{node priority} by assigning a priority value to every node in the program
and by introducing coordination facts that manipulate such priority values.
By default, nodes do not have any priority and can be picked in any order. In
our implementation, we use a FIFO approach because this tends to work better for
most programs since older nodes tend to a higher number of unexamined facts.

@@

LM provides the sensing fact \code{priority(node A, node B, float P)} in order to sense the
priority \texttt{P} of node \texttt{B} (at node \texttt{A}).
Sensing facts are only used in the body of the rules in order
to fetch information from the runtime system.
Note that when sensing facts are used to prove new facts, they are re-derived automatically.
~\footnote{The observant reader will know that all the coordination facts are
   linear and XXX
Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
value of \texttt{priority} facts, but this done automatically by the runtime
system.}

>>>>

In order to allow different scheduling strategies, we introduce the concept of
\emph{node priority} by assigning a priority value to every node in the program
and by introducing coordination facts that manipulate such priority values.
By default, nodes have no priority and can be picked in any order. In
our implementation, we use a FIFO approach because
older nodes tend to a higher number of unexamined facts, from which to derive
subsequent new facts.

@@

CLM provides the sensing fact \code{priority(node A, node B, float P)} in order to sense the
priority \texttt{P} of node \texttt{B} (at node \texttt{A}).
Sensing facts are only used in the body of the rules in order
to fetch information from the runtime system.
Note that when sensing facts are used to prove new facts, they are re-derived automatically.
~\footnote{The observant reader will notice that all the coordination facts are
linear and when created in the body of a rule will be consumed. The system
creates the necessary code to re-derive them without programmer interaction.
Likewise, \texttt{set-priority} and \texttt{set-default-priority} update the
value of \texttt{priority} facts by consuming and re-deriving them but this is
done automatically by the runtime system.}

### 2014-11-08 heat-transfer.tex
More updates.

In the Heat Transfer algorithm~(HT), we have a graph where heat values are
exchanged between nodes. The goal of the program is to exchange heat to a point
where new heat value $\delta = |H_i -
H_{i-1}| \le \epsilon$ for all nodes of the graph. The algorithm works
asynchronously since heat values can be updated by using new partial information
coming from neighbor nodes. This increases parallelism since nodes do not need
to synchronize between iterations.

Fig.~\ref{code:ht} shows the HT logical rules that send new heat values to
neighbor nodes. In the first rule we added \texttt{add-priority} to increase the priority of the neighbor
nodes if the current node has a large $\delta$. The idea is to prioritize the
computation of heat values of nodes (using \texttt{update}) that have a neighbor
that changed significantly. Multiple \texttt{add-priority} facts will
increase the priority of a node so that nodes with multiple large deltas will
have more priority. To further improve scalability, we added the third rule to
ignore to not send small $\delta$ values if the target node is in another
thread. Note that the LM compiler will merge the three rules in
Fig.~\ref{code:ht} into a single one that handles the 3 branches.

@@

new-heat(A, New, Old)
fabs(New - Old) <= epsilon
\underline[cpu-id(A, B, C)],
\underline[cpu-id(A, A, C)]
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New)}.

new-heat(A, New, Old)
fabs(New - Old) <= epsilon,
\underline[cpu-id(A, B, C1)],
\underline[cpu-id(A, A, C2)],
\underline[C1 <> C2]
   -o 1. // nothing is derived
\end{Verbatim}
  \caption{Coordination code for the Heat Transfer program. We updated the first rule
     to increase the priority of neighbor nodes and added the third rule to not
     send small $\delta$ if the neighbor is in another thread. The code logic remains exactly
     the same as before, however bigger changes in heat values are now
     propagated faster.}
  \label{code:ht}
\end{figure}
\normalsize

@@

>>>>

In the Heat Transfer algorithm~(HT), we have a graph where heat values are
exchanged between nodes. The goal of the program is to exchange heat between
nodes. The program stops when the new heat values of every node are
$\delta = |H_i - H_{i-1}| \le \epsilon$. The algorithm works
asynchronously since heat values can be updated by using new partial information
coming from neighbor nodes. This increases parallelism since nodes do not need
to synchronize between iterations.

Fig.~\ref{code:ht} shows the HT rules that send new heat values to
neighbor nodes. In the first rule we added \texttt{add-priority} to increase the priority of the neighbor
nodes if the current node has a large $\delta$. The idea is to prioritize the
computation of heat values of nodes (using \texttt{update}) that have a neighbor
that changed significantly. Multiple \texttt{add-priority} facts will
increase the priority of a node so that nodes with multiple large deltas will
have more priority.

@@

new-heat(A, New, Old)
fabs(New - Old) <= epsilon
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New)}.
\end{Verbatim}
  \caption{Coordination code for the Heat Transfer program. We updated the first rule
     to increase the priority of neighbor nodes.  The code logic remains exactly
     the same as before, however bigger changes in heat values are now
     propagated faster.}
  \label{code:ht}
\end{figure}
\normalsize

@@

To further improve locality, we split the second rule to
not send small $\delta$ values if the target node is in another
thread. XXX talk about results

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
new-heat(A, New, Old)
fabs(New - Old) <= epsilon
\underline[cpu-id(A, B, C)],
\underline[cpu-id(A, A, C)]
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New)}.

new-heat(A, New, Old)
fabs(New - Old) <= epsilon,
\underline[cpu-id(A, B, C1)],
\underline[cpu-id(A, A, C2)],
\underline[C1 <> C2]
   -o 1. // nothing is derived
\end{Verbatim}
  \caption{To improve locality, we added a third rule to not send small $\delta$
     values if the neighbor is in another thread.}
  \label{code:ht_better}
\end{figure}
\normalsize

### 2014-11-08 implementation.tex
More updates.

Coordination directives are compiled in two different ways, depending on whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of being added to the database as facts.
Semantically, action facts are like any other. However, since they are
immediatelly used by the machine, there is no need to store them in the
database, avoiding unnecessary allocations and deallocations.

@@

The virtual machine is implemented in C++11 and uses the threading system from
the standard library to implement multithreading. Each thread is responsible for
executing a subset of \emph{active nodes}. A node is active if it has unexamined
facts. After a node is processed, it becomes inactive until a new fact is
derived for it. Inactive nodes have no work to do. When a new fact is derived for an
inactive node, the node is made active and placed on the appropriate queue of
the appropriate thread.  Threads do useful work by processing active nodes from their queues. Whenever a
thread becomes idle, it attempts to steal nodes from a random thread.
If unsucessful, the thread becomes idle and waits for program
termination while periodically attempting to steal nodes.

@@

Threads pop active nodes from their queues and execute new candidate rules. 
In order to reduce inter-thread communication, node priorities are implemented
at the thread level. This means that, when a thread picks the highest
priority node, it is actually the highest priority from the set of nodes owned
by the thread and not the highest priority node in the whole program.
To allow node priorities, we use 2 two pairs of queues per thread: a pair of doubly linked
lists known as the \emph{standard queue} and a pair of min/max heap known as the \emph{priority queue}.

@@

In the paper~\cite{cruz-ppdp14}, the LM virtual machine was measured and
compared against implementations of similar programs but written in other
languages. For instance, the LBP program was found to be around 2 times slower
than GraphLab, while N Queens program was found to be around 10 to 15 slower
than a sequential C implementation. Some of those programs were compared against
a Python implementations and LM faired fairly well. In this section, we measure
the impact of the coordination mechanisms we have implemented for the new
virtual machine.

>>>>

Coordination directives are compiled in two different ways, depending on whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of being added to the database as facts.
Semantically, action facts are like any other. However, since they are
immediatelly used by the machine, there is no need to store them in the
database, therefore avoiding unnecessary allocations and deallocations.

@@

The virtual machine is implemented in C++11 and uses the threading system from
the standard library to implement multithreading. Each thread is responsible for
executing a subset of \emph{active nodes}. A node is active if it has unexamined
facts. After a node is processed, it becomes inactive until a new fact is
derived for it. When a new fact is derived for an
inactive node, the node is made active and placed on the appropriate queue of
the appropriate thread.  Threads do useful work by processing active nodes from their queues. Whenever a
thread becomes idle, it attempts to steal nodes from a random thread.
If unsucessful, the thread becomes idle and waits for program
termination while periodically attempting to steal nodes.

@@

Threads pop active nodes from a thread local queue and execute new candidate rules to
generate new facts, which in turn can activate other nodes.
In order to reduce inter-thread communication, node priorities are implemented
at the thread level. This means that, when a thread picks the highest
priority node, it is actually the highest priority from the set of nodes owned
by the thread and not the highest priority node in the whole program.
To allow node priorities, we use 2 two pairs of queues per thread: a pair of doubly linked
lists known as the \emph{standard queue} and a pair of min/max heap known as the \emph{priority queue}.

@@

In the paper~\cite{cruz-ppdp14}, the LM virtual machine was measured and
compared against implementations of similar programs but written in other
languages. For instance, the LBP program was found to be around 2 times slower
than GraphLab, while N Queens program was found to be around 10 to 15 slower
than a sequential C implementation. Some of those programs were compared against
a Python implementations and LM faired fairly well. In this section, we measure
the impact of the coordination mechanisms we have implemented for the CLM
virtual machine.

### 2014-11-08 minimax.tex
More updates.

The algorithm proceeds by building a game tree, where each tree node represents a
game state and the children represent the possible game moves that can
be made by either player 1 or player 2.
An evaluation function \texttt{f(State)} is used to compute the score of
the board for each leaf of the tree. A node is a leaf when the game state can no longer be
expanded. Finally, the
algorithm recursively minimizes or maximizes the scores of each node.
To select the best move for player 1, the
algorithm picks the move maximized at the root node.

In LM, the program starts with a root node (with the initial game state)
that is expanded with the available moves at each level. The graph of the
program is dynamic since nodes are created and then deleted once they are no
longer needed. The latter happens when the
leaf scores are computed or when a node fully minimizes or maximizes the
children scores. When the program ends, only the root node has facts in its
database.

The code in Fig.~\ref{minimax:check-end} shows part of the program.
The first three rules (lines 1-10) deal
with the case where no children nodes are created and the last three rules
(12-29) deal with the cases for creating new nodes.
The rule in lines 12-25 generates new nodes using the
\texttt{exists} language construct, which creates a new node address
\texttt{B}, where a new \texttt{play} fact is created with the a
board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
communicate with its parent.

As noted before in Section~\ref{sec:fifo}, the runtime schedules nodes by XXX,
in a FIFO ordering for nodes,
meaning that when children are expanded, we first expand the children and
then the children of the children. In this case, the defined ordering
is not optimal for since the complete tree needs
to be expanded before computing the scores at the leaves. The memory
complexity of the program is $\mathcal{O}(n)$, where $n$ is the number of nodes
in the tree.

@@

With coordination, we set the priority nodes to be the same as the
depth so that the tree is expanded in a depth-first fashion (lines 15 and 22).
The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads. As an example,
consider a system with 2 threads, $T_1$ and $T_2$, where $T_1$ first expands the root
node and then the first child. Since $T_2$ is idle, it steals half of the root's
children nodes and starts expanding one of the nodes in a depth-first fashion.
Since threads prioritize deeper nodes, the scores of the first leaves are immediatelly
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

@@

In Fig.~\ref{results:memory_minmax} we compare the memory usage and scalability
of the coordinated MiniMax against the regular MiniMax. The coordinated version
sees a huge reduction of memory usage, needing at most 8MB for 16 threads, while
the regular version needs almost 14GB. Note that as the number of threads goes
up, memory usage also goes up. This is an artifact of our parallel memory
allocator that allocates large chunks of memory beforehand.
In terms of scalability, our experimental results show a 14-fold speedup for the
coordinated version against a 12-fold for the regular version when using 16
threads. When comparing the two versions directly, there is a 20\% run time
reduction when using 16 threads in the coordinated version.

>>>>

The algorithm proceeds by building a game tree, where each tree node represents a
game state and the children represent the possible game moves that can
be made by either player 1 or player 2.
An evaluation function is used to compute the score of
the board for each leaf of the tree. A node is a leaf when the game state can no longer be
expanded. Finally, the
algorithm recursively minimizes or maximizes the scores of each node.
To select the best move for player 1, the
algorithm picks the move maximized at the root node.

In CLM, the program starts with a root node (with the initial game state)
that is expanded with the available moves at each level. The graph of the
program is dynamic since nodes are created and then deleted once they are no
longer needed. The latter happens when the
leaf scores are computed or when a node fully minimizes or maximizes the
children scores. When the program ends, only the root node has facts in its
database.

The code in Fig.~\ref{minimax:check-end} shows deals with expanding the tree.
The first three rules (lines 1-10) deal
with the case where no children nodes are created and the last three rules
(12-29) deal with the cases for creating new nodes.
The rule in lines 12-25 generates new nodes using the
\texttt{exists} language construct, which creates a child node 
\texttt{B}. We link \texttt{B} with its parent (\texttt{parent(B, A)})
and kick start the expansion of that node \texttt{B} by adding a \texttt{play}
fact.

As noted before in Section~\ref{sec:fifo}, the runtime schedules nodes by
default, which requires a complete expansion of the tree before computing the
scores at the leaves, leading
to a memory complexity of $\mathcal{O}(n)$, where $n$ is the number of nodes in
the tree.

@@

With coordination, we set the priority of a node to be its depth
so the tree is expanded in a depth-first fashion (lines 15 and 22),
leading to a memory complexity of $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads.
Since threads prioritize deeper nodes, the scores of the first leaves are immediatelly
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

\iffalse
As an example,
consider a system with 2 threads, $T_1$ and $T_2$, where $T_1$ first expands the root
node and then the first child. Since $T_2$ is idle, it steals half of the root's
children nodes and starts expanding one of the nodes in a depth-first fashion.
\fi

@@

In Fig.~\ref{results:memory_minmax} we compare the memory usage and scalability
of the coordinated MiniMax against the regular MiniMax. The coordinated version
uses significantly less memory (at most 8MB for 16 threads) than the regular
version (almost 14GB). Note that as the number of threads goes
up, memory usage also goes up. This is an artifact of our parallel memory
allocator that allocates large chunks of memory beforehand.
In terms of scalability, our experimental results show a 14-fold speedup for the
coordinated version against a 12-fold for the regular version when using 16
threads. When comparing the two versions directly, there is a 20\% run time
reduction when using 16 threads in the coordinated version.

### 2014-11-08 nqueens.tex
More updates.

The LM solution considers the chess board as a graph where nodes exchange valid
configurations between each other. First, we start with the empty
state in the first row of the chess board. Then, each square adds its own
position to the state and sends the state down, to the next row. Once a square
receives new configurations, it attempts to add its position to the
configurations. If valid, that configuration is then sent, recursively, to the
next row, until all rows are traversed. At the end of the program, the nodes at
the bottom row will have all the valid configurations.

Since computation goes from the top row to the bottom row, not all placements of
nodes on threads may be equally performant. This is specially true because the
bottom rows tend to perform the most work. The best placement is then to split the
board vertically, so that, in optimal conditions, each thread gets the same
number of columns.  Our experiments show that, on shared memory, it does not
matter much if we start with a bad placement since node stealing overcomes those
issues by improving load balancing dynamically. But what if we want run the
program on a distributed system, where moving nodes between workers can be
expensive? We measured the performance of the program without node stealing in
order to assess how important data placement would be for such systems.
We used the default partition of the program as it is, which results in the board being
split by rows and another configuration using the coordination action
\texttt{set-cpu(A, vertical(X, Y))} that forces nodes to be moved to another
thread using a vertical partitioning. Note that \texttt{X} and \texttt{Y} are
the coordinates of the node in the chessboard.

>>>>

The CLM solution considers the chess board as a graph where nodes exchange valid
configurations between each other. First, we start with the empty
state in the first row of the chess board. Then, each square adds its own
position to the state and sends the state down, to the next row. Once a square
receives new configurations, it attempts to add its position to the
configurations. If valid, that configuration is then sent, recursively, to the
next row, until all rows are traversed. At the end of the program, the nodes at
the bottom row will have all the valid configurations.

Since computation goes from the top row to the bottom row, not all placements of
nodes on threads may be equally performant. This is especially true because the
bottom rows tend to perform the most work. The best placement is then to split the
board vertically, so that, in optimal conditions, each thread gets the same
number of columns.  Our experiments show that, on shared memory, it does not
matter much if we start with a bad placement since node stealing overcomes those
issues by improving load balancing dynamically. But what if we want run the
program on a distributed system, where moving nodes between workers can be
expensive? We measured the performance of the program without node stealing in
order to assess how important data placement would be for such systems.
We used the default partition of the program as it is, which results in the board being
split by rows and another configuration using the coordination action
\texttt{set-cpu(A, vertical(X, Y))} that forces nodes to be moved to another
thread using a vertical partitioning. Note that \texttt{X} and \texttt{Y} are
the coordinates of the node in the chessboard.

### 2014-11-08 splash-bp.tex
More updates.

Randomized and approximation algorithms can obtain significant benefits from
coordination directives because although the final program results will not be
exact, they follow important statistical properties and can be computed faster.
Examples of such programs are Loopy Belief Propagation.
Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
where nodes exchange messages with their immediate neighbors and apply some computations to the messages
received.

LBP is an algorithm that maps very well to the graph based model of LM. In its
original form, the belief values of nodes are computed by synchronous iterations.
LBP offers more concurrency when belief values are computed asynchronously
leading to faster convergence. For this, every node keeps track of all messages
sent/received and recomputes the belief using partial information from neighbor
nodes. It is then possible to prioritize the computation of beliefs when a
neighbor's belief changes significantly.

The asynchronous approach proves to be a nice improvement over the synchronous
version. Still, it is possible to do even better. Gonzalez et
al~\cite{Gonzalez+al:aistats09paraml} developed an optimal algorithm to compute
this algorithm by first building a tree and then updating the beliefs of each
node twice, first from the leaves to the root and then from the root to the
leaves. The root of this tree is the node with the highest priority (based on
belief) while the other nodes in the tree must have a non-zero priority.
Note that the priorities are updated whenever a neighbor updates their belief.
These \emph{splash trees} are built iteratively until we reach convergence.
In Fig.~\ref{splash_bp} we represent two threads creating two different splash
trees. It is encouraged to create several splash trees concurrently as long as
we have threads to create them.

\begin{figure}[h!]
   \begin{center}
      \includegraphics[width=6.5cm]{figures/splash_bp}
   \end{center}
   \caption{Creating splash trees for belief propagation. Each threads picks the
   highest priority node and creates a tree from that node. The belief values
   are updated in two phases: first from the leaves to the root and then from
   the root to the leaves.}
   \label{splash_bp}
\end{figure}

The code for Splash Belief Propagation~(SBP) in Fig.~\ref{code:sbp} presents the
coordination code for LBP.  Please note that we just appended the code in
Fig.~\ref{code:sbp} to a working but unoptimized version of the algorithm, every
other logical rule remains the same. We add new logical ruels that coordinate
the creation and execution of the splash trees:

@@

Note that the \texttt{set-static} and \texttt{set-cpu} action fact are used in
line 2 in order to (1) force nodes to stay in the thread and (2) to partition
nodes as a grid of threads. This setups a well defined area of nodes
for threads to build splash trees on.

@@

// SECOND PHASE
second-phase(A, [], _)
   -o *underline{set-priority(A, 0.0)}, inactive(A), update(A).
second-phase(A, [A], Back)
   -o update(A), inactive(Back),
      inactive(A), *underline{set-priority(A, 0.0)}.
second-phase(A, [A, B | Next], Back)
   -o update(A), inactive(Back), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
\end{Verbatim}
  \caption{Coordination code for the Splash Belief Propagation program. LM needs
     50 lines of rules to implement splash trees, while GraphLab needs a
     total of 350 lines of C++ to implement the same functionality.  Note
     that when a linear fact is prefixed by \texttt{@}, it indicates that the
     fact is going to be re-derived.}
  \label{code:sbp}
\end{figure}
\normalsize

\begin{figure*}[ht!]
   \begin{center}
      \subfloat[]{\includegraphics[width=5cm]{results/system_belief-propagation-400.png}}
      \subfloat[]{\includegraphics[width=5cm]{results/system_splash-bp-400.png}}
      \subfloat[]{\includegraphics[width=5cm]{results/system_improve_belief-propagation-400.png}}
   \end{center}
   \caption{Experimental results for LBP and SBP. Figure (a) shows the
      scalability of LBP for both LM and GraphLab and Figure (b) shows the
      scalability of SBP. Figure (c) presents the
      improvements seen in SBP against LBP, where SBP runs, on average, 1.5 to 2.5
      times faster than LBP.}
   \label{results:splash_bp}
\end{figure*}

In this program, coordination assumes a far more important role than we have
seen before. Coordination rules fully drive the behavior of the algorithm and
although the final result of the algorithm is identical to the original
algorithm (minus probabilitistic errors), SBP works in a very different way.
A system that also implements SBP is
GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine algorithms.
GraphLab provides the splash scheduler for these types of inferences as part of
the framework. This particular scheduler is around 350 lines of complicated C++ code.
With our coordination facts, it is possible to create fairly complicated coordination
patterns with only 12 simple logical rules.

We measured the behavior of LBP and SBP for both LM and GraphLab.
Fig.~\ref{results:splash_bp} shows that both systems have very similar behavior
when using a variable number of threads.
Note that GraphLab avoids message passing between nodes since new neighbor
belief updates are changed immediatelly, resulting in slightly better
scalability for GraphLab.

>>>>

Randomized and approximation algorithms can obtain significant benefits from
coordination directives because their inherent non-determinism can be harnessed
to evaluate rules in different orders.
An example of such program is Loopy Belief Propagation.
Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
where nodes exchange messages with their immediate neighbors and apply some computations to the messages
received.

LBP is an algorithm that maps very well to the graph based model of CLM. In its
original form, the belief values of nodes are computed by synchronous iterations.
LBP offers more concurrency when belief values are computed asynchronously
leading to faster convergence. For this, every node keeps track of all messages
sent/received and recomputes the belief using partial information from neighbor
nodes. It is then possible to prioritize the computation of beliefs when a
neighbor's belief changes significantly.

The asynchronous approach proves to be a nice improvement over the synchronous
version. Still, it is possible to do even better. Gonzalez et
al~\cite{Gonzalez+al:aistats09paraml} developed an optimal algorithm to compute
this algorithm by first building a tree and then updating the beliefs of each
node twice, first from the leaves to the root and then from the root to the
leaves. The root of this tree is the node with the highest priority (based on
belief) while the other nodes in the tree must have a non-zero priority.
Note that the priorities are updated whenever a neighbor updates their belief.
These \emph{splash trees} are built iteratively until we reach convergence.
%In Fig.~\ref{splash_bp} we represent two threads creating two different splash trees.

\iffalse
\begin{figure}[h!]
   \begin{center}
      \includegraphics[width=6.5cm]{figures/splash_bp}
   \end{center}
   \caption{Creating splash trees for belief propagation. Each threads picks the
   highest priority node and creates a tree from that node. The belief values
   are updated in two phases: first from the leaves to the root and then from
   the root to the leaves.}
   \label{splash_bp}
\end{figure}
\fi

The code for Splash Belief Propagation~(SBP) in Fig.~\ref{code:sbp} presents the
coordination code for LBP.  Please note that we just appended the code in
Fig.~\ref{code:sbp} to a working but unoptimized version of the algorithm, every
other rule remains the same. We add new rules that coordinate
the creation and execution of the splash trees:

@@

Note that the \texttt{set-static} and \texttt{set-cpu} action fact are used in
line 2 in order to (1) force nodes to stay in the thread and (2) to partition
nodes as a grid of threads. This sets up a well defined area of nodes
for threads to build splash trees on.

@@

// SECOND PHASE
second-phase(A, [], _)
   -o *underline{set-priority(A, 0.0)}, inactive(A), update(A).
second-phase(A, [A], Back)
   -o update(A), inactive(Back),
      inactive(A), *underline{set-priority(A, 0.0)}.
second-phase(A, [A, B | Next], Back)
   -o update(A), inactive(Back), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
\end{Verbatim}
  \caption{Coordination code for the Splash Belief Propagation program. CLM needs
     50 lines of rules to implement splash trees, while GraphLab needs a
     total of 350 lines of C++ to implement the same functionality.  Note
     that when a linear fact is prefixed by \texttt{@}, it indicates that the
     fact is going to be re-derived.}
  \label{code:sbp}
\end{figure}
\normalsize

\begin{figure*}[ht!]
   \begin{center}
      \subfloat[]{\includegraphics[width=5cm]{results/system_belief-propagation-400.png}}
      \subfloat[]{\includegraphics[width=5cm]{results/system_splash-bp-400.png}}
      \subfloat[]{\includegraphics[width=5cm]{results/system_improve_belief-propagation-400.png}}
   \end{center}
   \caption{Experimental results for LBP and SBP. Figure (a) shows the
      scalability of LBP for both CLM and GraphLab and Figure (b) shows the
      scalability of SBP. Figure (c) presents the
      improvements seen in SBP against LBP, where SBP runs, on average, 1.5 to 2.5
      times faster than LBP.}
   \label{results:splash_bp}
\end{figure*}

In this program, coordination assumes a far more important role than we have
seen before. Coordination rules fully drive the behavior of the algorithm and
although the final result of the algorithm is identical to the original
algorithm (minus probabilitistic errors), SBP works in a very different way.
SBP is implemented in GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine algorithms.
GraphLab provides the splash scheduler for these types of inferences as part of
the framework. This particular scheduler is around 350 lines of C++ code.
With our coordination facts, it is possible to create the necessary
patterns with only 12 simple rules.

We measured the behavior of LBP and SBP for both CLM and GraphLab.
Fig.~\ref{results:splash_bp} shows that both systems have very similar behavior
when using a variable number of threads.  The differences in performance between
GraphLab and CLM comes mainly because GraphLab performs in-place updates of
neighbor belief values, while CLM compiler and runtime system performs naive
manipulation of facts by deriving rules. The in-place update could be generated
by the compiler using smarter compilation strategies.

### 2014-11-08 sssp.tex
More updates.

In Section~\ref{sec:language}, we said that the SSSP program has an optimal
scheduling if the nodes with the shortest distances are selected to run before
other nodes.  The coordinated version of the SSSP
program is presented in Fig.~\ref{code:shortest_path_program_coord} and it
uses one coordination fact, namely \texttt{set-priority} (line 14). Note that we
also use a global program directive to order priorities in ascending order (line
5).

When using a single thread, the algorithm behaves like
Dijkstra's shortest path algorithm~\cite{Dijkstra}. However, when using multiple
threads, each thread will pick the smallest distance from their subset of nodes.
While this does not yield the optimal program with relation to 1 thread, it
allows for parallel execution and locally avoid unnecessary work. XXX

@@

Figure~\ref{results:sssp_uspowergrid} shows experimental results a
SSSP program that computes the SSSP for 20\% of the nodes of the graph.
There are some situations where unecessary facts are propagated
because although the shortest distance is selected, sub-optimal distances may be
propagated because many SSSP distances are computed at the same time.
However, we see a reduction of over 50\% in the number of
derived facts when using the coordinated version \textbf{C} over the regular
version \textbf{R}. The results also show that the cordinated
version using 16 threads is 1.3 times faster than the regular version using 16
threads. In turn, this makes the coordinated version 20 times faster
than the regular version using 1 thread. Note that the coordinated program is as
scalable as the regular program.

>>>>

Here we add coordination to the SSSP program described in Section
The coordinated version of the
SSSP~(Fig.~\ref{code:shortest_path_program_coord}) uses the coordination fact
\texttt{set-priority} (line 14). We also use a global program directive to order
priorities in ascending order (line 5).

When run with one thread, the algorithm behaves like
Dijkstra's shortest path algorithm~\cite{Dijkstra}. When using multiple
threads, each thread will pick the smallest distance from their subset of nodes.
While this does not yield the optimal program with relation to 1 thread, it
allows for parallel execution and locally avoids unnecessary work. The result
scales well and it is close to Dijkstra's algorithm.

@@

Figure~\ref{results:sssp_uspowergrid} shows experimental results when the
program computes the SSSP for 20\% of the nodes of the graph.
There are some situations where unecessary facts are propagated
because although the shortest distance is selected locally, sub-optimal distances may be
propagated because many SSSP distances are computed at the same time.
However, we see a reduction of over 50\% in the number of
derived facts when using the coordinated version \textbf{C} over the regular
version \textbf{R}. The results also show that the cordinated
version using 16 threads is 1.3 times faster than the regular version using 16
threads while still showing good scaling.
